---
title: "A1_54007301"
author: "540007301"
date: today 
date-format: "DD MMMM, YYYY"
format:
  html:
    code-tools: true
    code-fold: true
    fig_caption: yes
    number_sections: yes
    embed-resources: true
    theme: flatly
    toc: true
    toc_depth: 4
    toc_float: true
execute:
  echo: true
  tidy: true
---

```{r loadpackages, message=FALSE, warning=FALSE}
library(Biobase)
library(limma)
library(kableExtra)
library(ggplot2)
library(caret)
library(randomForest)
library(pROC)
```

# Introduction.

This report investigates whether blood-based (GSE46474) or biopsy-based (GSE138043) gene expression samples lead to better classification of kidney graft rejection. Part 1 identifies the top differentially expressed genes in each dataset using limma. Part 2 visualises all features via volcano plots. Part 3 builds Random Forest classifiers evaluated with LOOCV, and compares their performance to determine which sample type is more suitable for clinical deployment.

# Part 1: Exploratory Analysis

## Data pre-processing

```{r}
load("GSE138043.RData")
gse138043 <- gse

load("GSE46474.RData")
gse46474 <- gse

expr1 <- exprs(gse138043)
pd1 <- pData(gse138043)
fd1 <- fData(gse138043)

expr2 <- exprs(gse46474)
pd2 <- pData(gse46474)
fd2 <- fData(gse46474)
```

This section first conducts pre-processing on the existing data by - formatting nas and empty inputs as NA_character - conduct name/symbol extraction by splitting certain specific columns when "///" or "//" were used. Furthermore, for cases where multiple probes reacted, that section would be disregarded, as these cases indicate that it is ambiguous which probe has reacted, and thus means that the result is not trustworthy.

```{r}
extract_symbols_fd1 <- function(x) {
  if (is.na(x) || x == "") return(character(0))
  
  parts <- unlist(strsplit(x, " /// "))
  
  syms <- vapply(parts, function(part) {
    fields <- unlist(strsplit(part, " // "))
    if (length(fields) >= 2) fields[2] else NA_character_
  }, character(1))
  
  syms <- unique(na.omit(syms))
  syms <- syms[syms != ""]
  syms
}

extract_unique_symbol_fd1 <- function(x) {
  syms <- extract_symbols_fd1(x)
  if (length(syms) == 1) syms else NA_character_
}

extract_symbols_fd2 <- function(x) {
  if (is.na(x) || x == "") return(character(0))
  
  parts <- unlist(strsplit(x, " /// "))
  parts <- trimws(parts)
  parts <- unique(parts)
  parts <- parts[parts != ""]
  parts
}


extract_unique_symbol_fd2 <- function(x) {
  syms <- extract_symbols_fd2(x)
  if (length(syms) == 1) syms else NA_character_
}
```

Then necessary data from biopsy-based samples (gse138043) and blood-based samples (gse46474) were extracted to from group1 and group2 respectively. group1 consists of the AR/non-AR result of of the biopsy-based samples whilst group2 consists the same but for blood-based samples.

```{r}
#| echo: false
boxplot(expr1, las = 2, main = "GSE138043 (biopsy-based)")
boxplot(expr2, las = 2, main = "GSE46474 (blood-based)")
```

Further, normality and variance was tested for both biopsy and blood based sample via a box plot, which seems to suggest somewhat of a skew and not perfect normality. However, as can be seen from below, both data are roughly log-scaled which suggest that the use of limma pipeline, which does not rely on strict normality, is sensible and thus we can proceed.

```{r}
summary(as.vector(expr1))
summary(as.vector(expr2))
```

## Finding the most differential genes

In the below process, it takes the pData of biopsy-based samples and differentiates them into AR and non-AR and designs a matrix using those data. Then it fits every gene in to a linear model and compares which of the genes has the largest fold change using makeContrast function. The final output ensures that the top 20 genes are all unique as one gene may be represented by multiple probes on a micro array.

```{r}
#gse138043

group1 <- factor(pd1[["rejection at 12months post transplant:ch1"]])
group1_clean <- factor(make.names(as.character(group1)))

design1 <- model.matrix(~ 0 + group1_clean)
colnames(design1) <- levels(group1_clean)

rfit1 <- lmFit(expr1, design1)
cont1 <- makeContrasts(AR - non.AR, levels = design1)
fit1 <- contrasts.fit(rfit1, cont1)
fit1 <- eBayes(fit1)

res1 <- topTable(fit1, number = Inf, adjust.method = "BH")

res1$GeneSymbol <- vapply(
  fd1[rownames(res1), "gene_assignment"],
  extract_unique_symbol_fd1,
  character(1)
)

res1_clean <- res1[!is.na(res1$GeneSymbol) & res1$GeneSymbol != "", ]

res1_clean <- res1_clean[order(res1_clean$adj.P.Val, -abs(res1_clean$logFC)), ]

res1_unique <- res1_clean[!duplicated(res1_clean$GeneSymbol), ]

top20_table1 <- res1_unique[1:20, c("GeneSymbol", "logFC", "adj.P.Val", "P.Value", "t", "B")]
top20_table1
```

The same process has been conducted on the blood-based samples as well.

```{r}
#gse46474 
group2 <- factor(pd2[["procedure status:ch1"]])
group2_clean <- factor(make.names(as.character(group2)))

design2 <- model.matrix(~ 0 + group2_clean)
colnames(design2) <- levels(group2_clean)

fit2 <- lmFit(expr2, design2)
cont2 <- makeContrasts(
  post.transplant.acute.rejection..AR. - post.transplant.non.rejection..NR.,
  levels = design2
)
fit2 <- contrasts.fit(fit2, cont2)
fit2 <- eBayes(fit2)

res2 <- topTable(fit2, number = Inf, adjust.method = "BH")

res2$GeneSymbol <- vapply(
  fd2[rownames(res2), "Gene Symbol"],
  extract_unique_symbol_fd2,
  character(1)
)
res2_clean <- res2[!is.na(res2$GeneSymbol) & res2$GeneSymbol != "", ]

res2_clean <- res2_clean[order(res2_clean$adj.P.Val, -abs(res2_clean$logFC)), ]

res2_unique <- res2_clean[!duplicated(res2_clean$GeneSymbol), ]

top20_table2 <- res2_unique[1:20, c("GeneSymbol", "logFC", "adj.P.Val", "P.Value", "t", "B")]
```

Below are the final results for biopsy and blood based samples respectively.

biopsy-based:

```{r}
#| echo: false
kable(top20_table1)
```

blood-based:

```{r}
#| echo: false
kable(top20_table2)
```

Further, by testing the overlap between blood based and biopsy based samples, it shows that there are no overlaps at all and thus the top 20 for each are completely different sets of genes.

```{r}
#| echo: false
top20_1 <- top20_table1$GeneSymbol
top20_2 <- top20_table2$GeneSymbol
overlap_genes <- intersect(top20_1, top20_2)
biopsy_only    <- setdiff(top20_1, top20_2)
blood_only     <- setdiff(top20_2, top20_1)
```

# Part 2. Visualization

Volcano plots are used below to visualise all probes simultaneously for both the biopsy-based (p1) and blood-based (p2) datasets. The x-axis shows the log2 fold change (magnitude of differential expression) and the y-axis shows the -log10 p-value (statistical significance). Genes in the upper-right and upper-left corners are both highly significant and strongly differentially expressed, making them the most relevant candidates. This visualisation was chosen because it efficiently displays both effect size and significance for every feature in a single plot, allowing quick identification of potentially important genes.

```{r}
p1 <- ggplot(res1, aes(logFC,-log10(P.Value)))+
    geom_point(aes(colour=-log10(P.Value)), alpha=1/3, size=1) +
    scale_colour_gradient(low="blue",high="red")+
    xlab("log2 fold change") + ylab("-log10 p-value") +
    ggtitle("GSE138043 (Biopsy-based)")
p2 <- ggplot(res2, aes(logFC,-log10(P.Value)))+
    geom_point(aes(colour=-log10(P.Value)), alpha=1/3, size=1) +
    scale_colour_gradient(low="blue",high="red")+
    xlab("log2 fold change") + ylab("-log10 p-value") +
    ggtitle("GSE46474 (Blood-based)")
p1
p2
```

Taking a look at the graph above, it can be seen that in general, biopsy based blood samples tends to have lower magnitude of change compared to blood based samples, and the blood based samples seem to have changes that are more significant.

# Part 3: Model + Evaluation

The first is the model for the biopsy based samples, put into a random forest classification model. The model was chosen aboves the likes of k-means/kNN models due to the data set's nature of having high dimension and is robust to multicollinearity. Further, (Leave-one-out-cross-validation)LOOCV a specific kind of cross-validation technique was used to prevent over fitting. LOOCV, unlike a usual k-fold CV only leaves one sample out when validating, making the process more suitable for a relatively small samples size data, like the ones dealt in this case. By going through LOOCV process, the data is capable of testing on an "unknown" data, which the model in real life will be used on.

```{r}
set.seed(42)
actual_classes1 <- factor(make.names(as.character(pd1[["rejection at 12months post transplant:ch1"]])))
n_samples <- ncol(expr1)
predictions1 <- character(n_samples)
prob_scores1 <- numeric(n_samples)
positive_class1 <- levels(actual_classes1)[1]  # "AR"
for (i in 1:n_samples) {
  train_expr <- expr1[, -i]
  test_expr <- expr1[, i, drop = FALSE]
  
  train_group <- actual_classes1[-i]
  test_group <- actual_classes1[i]
  
  design_train <- model.matrix(~ 0 + train_group)
  colnames(design_train) <- levels(train_group)
  
  fit_train <- lmFit(train_expr, design_train)
  cont_train <- makeContrasts(AR - non.AR, levels = design_train) 
  fit_train <- contrasts.fit(fit_train, cont_train)
  fit_train <- eBayes(fit_train)
  
  res_train <- topTable(fit_train, number = Inf, adjust.method = "BH")
  res_train$GeneSymbol <- vapply(
    fd1[rownames(res_train), "gene_assignment"],
    extract_unique_symbol_fd1,
    character(1)
  )
  res_train_clean <- res_train[!is.na(res_train$GeneSymbol) & res_train$GeneSymbol != "", ]
  res_train_clean <- res_train_clean[order(res_train_clean$adj.P.Val, -abs(res_train_clean$logFC)), ]
  res_train_unique <- res_train_clean[!duplicated(res_train_clean$GeneSymbol), ]

  top20_probes_fold <- head(rownames(res_train_unique), 20)

  train_rf_data <- data.frame(t(train_expr[top20_probes_fold, ]), Status = train_group)
  colnames(train_rf_data) <- make.names(colnames(train_rf_data))

  test_rf_data <- data.frame(t(test_expr[top20_probes_fold, , drop = FALSE]))
  colnames(test_rf_data) <- make.names(colnames(test_rf_data))

  rf_model <- randomForest(Status ~ ., data = train_rf_data, ntree = 500)

  fold_prediction <- predict(rf_model, newdata = test_rf_data)
  predictions1[i] <- as.character(fold_prediction)

  fold_prob <- predict(rf_model, newdata = test_rf_data, type = "vote")
  prob_scores1[i] <- fold_prob[1, positive_class1]
}

predictions1 <- factor(predictions1, levels = levels(actual_classes1))
```

An identical model was used for blood-based samples as well, using random forest and LOOCV.

```{r}
set.seed(42)
actual_classes2 <- factor(make.names(as.character(pd2[["procedure status:ch1"]])))
n_samples <- ncol(expr2)
predictions2 <- character(n_samples)
prob_scores2 <- numeric(n_samples)
positive_class2 <- levels(actual_classes2)[1]
for (i in 1:n_samples) {
  train_expr <- expr2[, -i]
  test_expr <- expr2[, i, drop = FALSE]
  
  train_group <- actual_classes2[-i]
  test_group <- actual_classes2[i]
  
  design_train <- model.matrix(~ 0 + train_group)
  colnames(design_train) <- levels(train_group)
  
  fit_train <- lmFit(train_expr, design_train)
  cont_train <- makeContrasts(
  post.transplant.acute.rejection..AR. - post.transplant.non.rejection..NR.,
  levels = design_train
  )
  fit_train <- contrasts.fit(fit_train, cont_train)
  fit_train <- eBayes(fit_train)
  
  res_train <- topTable(fit_train, number = Inf, adjust.method = "BH")
  
  res_train$GeneSymbol <- vapply(
    fd2[rownames(res_train), "Gene Symbol"],
    extract_unique_symbol_fd2,
    character(1)
  )
  
  res_train_clean <- res_train[!is.na(res_train$GeneSymbol) & res_train$GeneSymbol != "", ]
  res_train_clean <- res_train_clean[order(res_train_clean$adj.P.Val, -abs(res_train_clean$logFC)), ]
  res_train_unique <- res_train_clean[!duplicated(res_train_clean$GeneSymbol), ]

  top20_probes_fold <- head(rownames(res_train_unique), 20)

  train_rf_data <- data.frame(t(train_expr[top20_probes_fold, ]), Status = train_group)
  colnames(train_rf_data) <- make.names(colnames(train_rf_data))

  test_rf_data <- data.frame(t(test_expr[top20_probes_fold, , drop = FALSE]))
  colnames(test_rf_data) <- make.names(colnames(test_rf_data))

  rf_model <- randomForest(Status ~ ., data = train_rf_data, ntree = 500)

  fold_prediction <- predict(rf_model, newdata = test_rf_data)
  predictions2[i] <- as.character(fold_prediction)

  fold_prob <- predict(rf_model, newdata = test_rf_data, type = "vote")
  prob_scores2[i] <- fold_prob[1, positive_class2]
}

predictions2 <- factor(predictions2, levels = levels(actual_classes2))

```

Based on it, below are the results from each of the models.

Biopsy based model

```{r}
#| echo: false
confusionMatrix(predictions1, actual_classes1)
```

The Biopsy model shows a respectable overall accuracy at 82.69%. However the sensitivity data shows how it is rather weak in terms of trying to catch all AR cases. On the contrary, this model performs substantially well in find in the true negative (i.e. catching cases in non-AR cases). Furthermore, the kappa score of 55% indicates model achieves 55% of the possible improvement over random guessing. Whilst the low p-value indicates that it is statistically better than simply guessing, it leaves a lot of question regarding the real accuracy of the model. The CI suggest that a range from 70~92% is likely the real accuracy of the model.

blood based model

```{r}
#| echo: false
confusionMatrix(predictions2, actual_classes2)
```

Blood based model has a slightly lower overall accuracy at 80%. However, when looking into the details, the data is substantially more balanced with sensitivity and specificity. This means that the likelihood of the model predicting truth and false is equally at 80%. In terms of the kappa score, it shows a moderate improvement relative to biopsy based model, but is still has room for improvements. The CI suggest that a range from 74\~90% is likely the real accuracy of the model, which is a larger range compared to the biopsy model.

## ROC Curves

Below are the ROC curves for both models, which plot the true positive rate against the false positive rate at various classification thresholds. the arc shape on both indicates that both have high sensitivity + few false alarms, and that both models in this perspective is not severely different.

```{r message=FALSE, warning=FALSE}
#| echo: false
roc1 <- roc(actual_classes1, prob_scores1, levels = rev(levels(actual_classes1)))
roc2 <- roc(actual_classes2, prob_scores2, levels = rev(levels(actual_classes2)))

plot(roc1, col = "blue", main = "ROC Curves: Biopsy vs Blood Model",
     legacy.axes = TRUE, print.auc = TRUE, print.auc.y = 0.4)
plot(roc2, col = "red", add = TRUE,
     print.auc = TRUE, print.auc.y = 0.3)
legend("bottomright", legend = c("Biopsy-based", "Blood-based"),
       col = c("blue", "red"), lwd = 2)
```

# Conclusion

Overall, the blood-based model is likely more reliable than the biopsy-based model due to its more balanced sensitivity and specificity, though both have room for improvement. LOOCV was chosen because it maximises training data in each fold, which is appropriate for these small sample sizes. However, LOOCV estimates can have high variance, so the reported accuracies should be interpreted with caution. Increasing the sample size would help produce more stable and generalisable performance estimates.
